---
title: "Applying Cluster Analysis in R for Business Segmentation"
subtitle: Data Analysis for Decision-Making (SS 2026)
author:
   - name: "Saketh Srigiri"
     email: srigiri.saketh@stud.hs-fresenius.de
     affiliation: "Hochschule Fresenius University of Applied Sciences"
date: "16 June 2026"
date-format: long
format:
  html: default
  pdf: default
execute: 
  echo: true
  warning: false
  message: false
editor: visual
bibliography: references.bib
csl: apa.csl
---

# 1. Introduction

Organizations generate large volumes of customer data through transactions, online interactions, loyalty programs, and social media. However, this data becomes valuable only when meaningful patterns are identified. Cluster analysis, an unsupervised machine learning technique, groups similar observations to uncover natural customer segments and support data-driven decision-making [@Gan2007; @Hastie2009].

Business segmentation through clustering enables personalized marketing, customer relationship management, and efficient resource allocation across industries. Among clustering methods, K-means is widely used for its simplicity and efficiency, while R provides powerful tools for data pre-processing, visualization, and implementation, making it an ideal platform for customer segmentation and business analytics [@Hastie2009; @Wickham2023].

# 2. Business Applications of Customer Segmentation

| Industry | Segmentation Basis | Business Outcome |
|------------------------|------------------------|------------------------|
| Retail | Purchase frequency, spending, preferences | Personalized promotions |
| Banking | Income, transactions, risk profile | Product recommendations, fraud detection |
| Telecommunications | Usage patterns, churn probability | Improved customer retention |
| E-commerce | Browsing and purchase behavior | Product recommendations, ad optimization |

# 3. K-means Clustering Algorithm

K-means partitions observations into K clusters by minimizing within-clusters variation [@Hastie2009]. The algorithm follows these steps:

1.  Select the number of clusters (K).
2.  Randomly initialize centroids.
3.  Assign each observation to its nearest centroid.
4.  Recalculate cluster centroids.
5.  Repeat until assignments stabilize.

Choosing the correct value of K is critical. Common methods include: - Elbow Method

-   Silhouette Analysis

-   Gap Statistic

# 4. Implementing Cluster Analysis in R

The following example demonstrates customer segmentation using R.

**Step 1: Load Packages**

```{r}
library(tidyverse)
library(cluster)
library(factoextra)
```

**Step 2: Create Sample Data**

```{r}
customers <- data.frame(
  CustomerID = 1:15,
  AnnualIncome = c(15, 18, 25, 30, 35, 45, 50, 55, 60, 65, 70, 75, 80, 90, 100),
  SpendingScore = c(39, 45, 50, 60, 65, 70, 75, 40, 35, 30, 80, 85, 25, 20, 15)
)
```

**Step 3: Standardize Variables**

```{r}
scaled_data <- scale(customers[,2:3])
```

**Step 4: Run K-Means**

```{r}
set.seed(123)

km <- kmeans(
  scaled_data,
  centers = 4,
  nstart = 25
)
```

**Step 5: Visualize Clusters**

```{r}
#| fig-height: 2
#| fig-width: 4
fviz_cluster(
  km,
  data = scaled_data
)
```

The visualization reveals two distinct customer groups representing higher-income, lower-income, higher-spending, and lower-spending patterns.

# **5. Advantages and Limitations**

Advantages: Simple to implement, fast on large datasets, easy to interpret. Limitations: K must be chosen in advance, assumes spherical clusters.

# 6. Evaluating Cluster Quality

One of the most crucial stages in cluster analysis is figuring out the ideal number of clusters. While choosing too many clusters could lead to needless complexity and decreased interpretability, choosing too few clusters could merge different client groups. Cluster quality is assessed using a number of statistical methods.

### **6.1 Elbow Method**

The Elbow Method plots WCSS against K values — the optimal K appears where the curve bends [@Hastie2009]. If the elbow appears at K = 3, then three clusters may provide the best balance between simplicity and explanatory power.

```{r}
#| fig-height: 1.8
#| fig-width: 2
library(factoextra)

fviz_nbclust(
  scaled_data,
  kmeans,
  method = "wss"
)
```

### **6.2 Silhouette Analysis**

Silhouette scores range from -1 to 1, where values close to 1 indicate well-defined clusters [@Rousseeuw1987]. This method often provides a more reliable estimate of the optimal number of clusters than relying solely on visual inspection.

```{r}
#| fig-height: 1.8
#| fig-width: 2
fviz_nbclust(
  scaled_data,
  kmeans,
  method = "silhouette"
)
```

### **6.3 Business Interpretation of Clusters**

The true value of clustering lies not only in identifying groups but also in translating those groups into actionable business strategies. Managers can use these insights to allocate marketing budgets more effectively and improve customer retention.

| Customer Segment | Characteristics | Business Strategy |
|------------------------|------------------------|------------------------|
| Budget Customers | Low income, low spending | Offer discounts and promotional bundles |
| Cautios Savers | High income, low spending | Highlight value for money and cashback offers |
| Aspirational Customers | Low to moderate income, high spending | loyalty programs and installment payment options |
| Premium Customers | High income, high spending | VIP memberships and personalized services |

# 7. Conclusion

Cluster analysis is a powerful unsupervised learning technique that helps organizations identify hidden patterns in customer data. Using R and algorithms such as K-means, businesses can group customers based on demographics or purchasing behavior to support targeted marketing and better decision-making. Effective implementation requires proper data pre-processing, standardization, and evaluation using methods like the Elbow Method and Silhouette Analysis. When combined with business knowledge, cluster analysis provides valuable insights that improve customer satisfaction and enhance organizational competitiveness.

# Affidavit

::: {appearance="default" icon="false"}
I hereby affirm that this submitted paper was authored unaided and solely by me. Additionally, no other sources than those in the reference list were used. Parts of this paper, including tables and figures, that have been taken either verbatim or analogously from other works have in each case been properly cited with regard to their origin and authorship. This paper either in parts or in its entirety, be it in the same or similar form, has not been submitted to any other examination board and has not been published.

Checklist:

-   [x] The handout contains 3-5 pages of text.
-   [x] The submission contains the Quarto file of the handout.
-   [x] The submission contains the Quarto file of the presentation.
-   [x] The submission contains the HTML file of the handout.
-   [x] The submission contains the HTML file of the presentation.
-   [x] The submission contains the PDF file of the handout.
-   [x] The submission contains the PDF file of the presentation.
-   [x] The title page of the presentation and the handout contain personal details (name, email, matriculation number).
-   [x] The handout contains a bibliography, created using BibTeX with an APA citation style.
-   [x] Either the handout or the presentation contains R code that proofs the expertise in coding.
-   [x] The filled out Affidavit.
-   [x] The link to the presentation and the handout published on GitHub.

Saketh Srigiri, 16 June 2026, Cologne
:::

# References
